---
title: "exploring the data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# strains
the files *starins.csv* has a list of strains isolated from chemstats
```{r}
rm(list=ls())
library(tidyverse, quietly = TRUE)
library(cowplot, quietly = TRUE)
d <- read_csv("../data/strains.csv")
rbind(head(d,n=3), rep("...",ncol(d)),tail(d,n=3))
```

How many Isolates are there of each kind?  
```{r}
d%>%
      filter(trt!="A")%>% #remove ancestor
       group_by(trt,lim,time) %>%
   summarize(isolates = n()) %>%
      ggplot(.,aes(x=time, y=isolates))+
      geom_bar(stat = "identity")+
      facet_grid(trt~lim)+
      theme_cowplot()
```
```{r}
d%>%
      filter(trt!="A")%>% #remove ancestor
       group_by(trt,lim) %>%
   summarize(isolates = n()) 
      
```

So there are isolates from both phage and nom phage chemostats. For N-limitation, no-phage isolates are only from firat few days.

#Breseq  
First chunck of table
```{r}
d <- read.csv("../data/breseq.compare.csv", header = TRUE)
head(d,n=11)
```

Looking at the code organizing the mutation data in MLs .Rmd file. The data is processed to have a positionXstrain matrix. at one point this code chunk appears  
```{r, eval=FALSE}
# Remove rows with no predicted mutations  
class(breseq) = "numeric"  
breseq <- breseq[-c(which(rowSums(breseq)==35)),]     
```  
At this point the table is 36 strains X 209 loci. I am not sure why the 35 cutoff was used. There are no rows with 36 mutaations, but there are rows with 34. there are 3 rows in whic 35/36 strains have mutations. In all 3 cases the same strainsoes not have this mutation, That strain is 03N1S1 (day 3, N-limited, chemostat 1, no-phage). Looking at the breseq file it seems this strain does not have any mutations. So in fact, the N-limited no phage chemostats have only day1 isolates. However this strain has not been removed from the next step of analysis.

Looing at the data partitioning into the +/- phage treatment, I think there is a problem.
In this next chunk filtering is by a test applied to subset of table (`breseq.minus[,6:117]`) and applied to full table (`breseq.minus`).

I am also not sure why the columns `118:211` are ignored.

```{r, eval = TRUE}


load(file = "breseq.all.Rdata")

breseq.minus <- breseq.all[breseq.all$trt=="Ph-",]
breseq.minus <- breseq.minus[,-c(which(colSums(breseq.minus[,6:117]) ==0))]
breseq.minus.muts = colnames(breseq.minus[3:41])
breseq.plus <- breseq.all[breseq.all$trt=="Ph+",]
breseq.plus <- breseq.plus[,-c(which(colSums(breseq.plus[,6:117]) ==0))]
breseq.plus.muts = colnames(breseq.plus[5:89])

both <- intersect(breseq.minus.muts,breseq.plus.muts)
phage.minus.only <- setdiff(breseq.minus.muts,breseq.plus.muts)
phage.plus.only <- setdiff(breseq.plus.muts,breseq.minus.muts)
```

I will subset it my way and compare

```{r, eval = TRUE}
sums <- 
breseq.all%>%
      filter(trt=="Ph-")%>%
      select(6:ncol(breseq.all))%>%
      colSums()

breseq.minus2 <- breseq.all%>%
      filter(trt=="Ph-")%>%
      select(as.numeric(which(sums!=0)+5))

sums <- 
breseq.all%>%
      filter(trt=="Ph+")%>%
      select(6:ncol(breseq.all))%>%
      colSums()

breseq.plus2 <- breseq.all%>%
      filter(trt=="Ph+")%>%
      select(as.numeric(which(sums!=0)+5))
      

breseq.minus.muts2 = colnames(breseq.minus2)
breseq.plus.muts2 = colnames(breseq.plus2)

both2 <- intersect(breseq.minus.muts2,breseq.plus.muts2)
phage.minus.only2 <- setdiff(breseq.minus.muts2,breseq.plus.muts2)
phage.plus.only2 <- setdiff(breseq.plus.muts2,breseq.minus.muts2)      

all.muts <- breseq.all%>%
     select(6:ncol(breseq.all))%>%
     colnames()

df <- data.frame(loci=as.numeric(gsub(",", "",all.muts)))
df$both.ML <- ifelse(all.muts%in%both,1,NA)
df$both.DS <- ifelse(all.muts%in%both2,1,NA)
df$noPHI.ML <- ifelse(all.muts%in%phage.minus.only,1,NA)
df$noPHI.DS <- ifelse(all.muts%in%phage.minus.only2,1,NA)
df$PHI.ML <- ifelse(all.muts%in%phage.plus.only,1,NA)
df$PHI.DS <- ifelse(all.muts%in%phage.plus.only2,1,NA)


df%>%
      gather(key=category,value = mutation, 2:7)%>%
      ggplot(aes(x=loci, y=category))+
      geom_vline(xintercept = df$loci, color="grey")+
      geom_jitter(aes(color=factor(mutation)), width=0, height=0.2)+
      theme_cowplot(font_size = 10)+
      scale_color_discrete(na.value = NA)
      
```

```{r}

df2 <- data.frame(loci = df$loci, both=NA, noPHI=NA, PHI=NA)
df2$both <- ifelse(is.na(df$both.ML), "NO","ML")
df2$both[!is.na(df$both.DS)] <- ifelse(df2$both[!is.na(df$both.DS)]=='ML', "BOTH", "DS")
df2$noPHI <- ifelse(is.na(df$noPHI.ML),"NO", "ML")
df2$noPHI[!is.na(df$noPHI.DS)] <- ifelse(df2$noPHI[!is.na(df$noPHI.DS)]=='ML', "BOTH", "DS")
df2$PHI <- ifelse(is.na(df$PHI.ML),"NO", "ML")
df2$PHI[!is.na(df$phi.DS)] <- ifelse(df2$PHI[!is.na(df$PHI.DS)]=='ML', "BOTH", "DS")

p <- 
df2%>%
      gather(key=category,value = mutation, 2:4)%>%
      ggplot(aes(x=category, y=factor(loci)))+
      geom_tile(aes(fill=mutation))+
      theme_cowplot(font_size = 10)+
      scale_fill_viridis(discrete=TRUE)+
      coord_flip()
p
      ggsave2(p,filename = "loci.compared.jpg")


```

Tabulating the data
I want to make a table of the genes in which mutations were found only in phage-treatment.

```{r}
load ("breseq.Rdata")
breseq <-unlist(breseq)
#remove the mutations that are in all (see above)
breseq <- breseq[(loci%in%all.muts),]
loci <- row.names(breseq)
# breseq <- as.data.frame(breseq)
breseq$loci <- loci
rownames(breseq) <- c()
meta <-breseq[,c(1,38:41)]


phi.filter <- df$loci[is.na(df$PHI.DS)]
PHI.only <- breseq[as.numeric(gsub(",","",loci))%in%phi.filter,]

#break up and put back together
meta <- PHI.only[,c(1,38:41)]
PHI.only <-  PHI.only[,-c(1,38:41)]
# rownames(PHI.only) <- c()
PHI.only <- as.matrix.data.frame(PHI.only)
PHI.only <- `class<-`(PHI.only,"numeric")
PHI.only <- t(PHI.only)
strains <- read_csv("../data/strains.csv")
full_join(strains, as.data.frame(PHI.only), by())

```


