---
title: "Mutation by strain table "
output: html_notebook
editor_options: 
  chunk_output_type: console
---
My goal here is to organize the breseq data in a t table with genes as rows and the strains as columns. Table cells will contain nubers indicating the nuber of mutations observed.


```{r, message=FALSE}
rm(list=ls())
library(tidyverse, quietly = TRUE)
library(cowplot, quietly = TRUE)
library(viridis)
library(vegan)
```
## Loading strain data  
* trt: phage infection  
* lim: limiting nutrient  
* cID: chemostat number  
* time: sampling time in days  
* strain_ID: sample day + chemostat + strain number  

```{r, message=FALSE}
# table of strains and their source (time, treatment)
strains <- read_csv("../data/strains.csv")
# remove ansector
strains <- filter(strains,strain_ID!="WH7803")
strains
```  

# Mutation data  
I copy pasted the data found in _data/20140611_allcompare.html_ into excel and cleaned it up a bit. Re-importing it preserved some of the charachters that were lost in the file as it is found in _data/breseq.compare.csv_.  
* I removed the first line that indicates a complete genome deletion. This results in strain 03N1S1 not having any mutation. this caused a bunch of NA being put in cells upon import, so I put in 0s instead.   
* I cleaned all the special charechters out of the mutation column. the right pointing arrow I replaced with "sub", $\Delta$ was replaced by "d", and a bunch of spces that acted funny upon import were changed to regular spaces.  
* Cleaning the anotation data from special charecters is too much of a mess. Instead I downloaded a table of the genetic features of WH7803 from <ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCA/000/063/505/GCA_000063505.1_ASM6350v1/GCA_000063505.1_ASM6350v1_feature_table.txt.gz>. I verified that the positions are similar to what is in the breseq html file in a few loci. seems like a perfect fit. This too required some cleanin to remove the duplicated rows and some empty columns.  
* replaced empty cells, "?", $\Delta$  with "0"; replaced "100%" with 1.  

```{r, message=FALSE}
# Import data file
breseq <- read_csv("breseq_allcompare.csv")

# correct column names
colnames (breseq)<- gsub(".g","",colnames (breseq))

```

# Summarizing mutations by gene  

```{r}


# # crop out the annotation data
# breseq.annot  <-  
#   breseq%>%
#     select(position,mutation,annotation,gene,description)


#summarize mutations by gene
mXg <- 
  breseq%>%
    select(-position,-mutation,-annotation)%>%
    group_by(gene, description)%>%
    summarize_if(is.numeric, sum, na.rm = TRUE)
   
mXg$gene <-  str_replace_all(mXg$gene, "[^[:alnum:],_,/]", "")
# write_csv(mXg, "mxg.csv")
knitr::kable(mXg)

# Principal Coordinates Analysis (PCoA)

# Take count data and use gene names as rownames
mXg.1 <-as.data.frame(mXg[,3:ncol(mXg)])
rownames(mXg.1)<-t(mXg[,1])

# Subtract out mutations found in no-phage controls
# Probably a better way to assign isolate IDs to treatments
# Here is just did it manually by cross referencing Megan's excel sheet

N.no.phage.list <- c("01N2S3", "01N3S3", "01N5S4", "03N1S1")
N.no.phage.muts <- rowSums(mXg.1[, N.no.phage.list])
N.no.phage.PA <-(N.no.phage.muts >0)*1
N.plus.phage.list <- c("03N2S5","03N3S2","03N3S4","03N5S2",
                       "05N2S5","05N3S3","16N5S4","19N3S4","21N2S3","21N3S2")
N.plus.phage <- (mXg.1[, N.plus.phage.list] >0) *1
N.plus.phage.PA <- (N.plus.phage - N.no.phage.PA >0)*1

P.no.phage.list <- c("01P3S1","01P4S5","01P5S1","03P3S5","05P3S1",
                    "12P3S4","16P3S2","19P3S4","21P3S4")
P.no.phage.muts <- rowSums(mXg.1[, P.no.phage.list])
P.no.phage.PA <-(P.no.phage.muts >0)*1
P.plus.phage.list <- c("01P3S1","01P4S5","01P5S1","03P3S5","05P3S1",
                    "12P3S4","16P3S2","19P3S4","21P3S4")
P.plus.phage <- (mXg.1[, P.plus.phage.list] >0) *1
P.plus.phage.PA <- (P.plus.phage - P.no.phage.PA >0)*1

phage.muts.complete <- data.frame(N.plus.phage.PA,P.plus.phage.PA)
phage.muts <- phage.muts.complete[ rowSums(phage.muts.complete)!=0, ] 
colnames(phage.muts) <-c(colnames(N.plus.phage.PA),colnames(P.plus.phage.PA))

# Transpose sample x mutation matrix, calculate distance matrix, run PCoA
# There are a couple of errors called with vegdist and cmdscale
#mXg.2 <- t(mXg.1)
mXg.2 <- t(phage.muts)
mut.bc <- vegdist(mXg.2, method = "bray", binary = "FALSE") #bray, jaccard, canbera
mut.PCoA <- cmdscale(mut.bc, eig = TRUE, k = 3)
explainvar1 <- round(mut.PCoA$eig[1] / sum(mut.PCoA$eig), 3) * 100
explainvar2 <- round(mut.PCoA$eig[2] / sum(mut.PCoA$eig), 3) * 100
explainvar3 <- round(mut.PCoA$eig[3] / sum(mut.PCoA$eig), 3) * 100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

# Plot PCoA
plot.new()
par(mar = c(4, 5, 4, 0) + 0.5)

plot(mut.PCoA$points[ ,1], mut.PCoA$points[ ,2],
     ylim = c(-0.4, 0.4), xlim = c(-0.2, 0.2),
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
     #xlab = "", ylab = "", xaxt = "n", yaxt = "n",
     pch = 22, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1,
     axes = FALSE)

# Add Axes
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.25, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.25, las = 1)
axis(side = 3, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=-0.02)
axis(side = 4, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=-0.02)
axis(side = 1, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=0.01)
axis(side = 2, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=0.01)
axis(side = 3, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=0.01)
axis(side = 4, labels = F, lwd.ticks = 2, cex.axis = 1, las = 1, tck=0.01)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

# Subset data for plotting

#N.no.phage.list <- c("01N2S3", "01N3S3", "01N5S4", "03N1S1")
#N.no.phage<-mut.PCoA$points[N.no.phage.list, ]

N.plus.phage.list <- c("03N2S5","03N3S2","03N3S4","03N5S2",
                       "05N2S5","05N3S3","16N5S4","19N3S4","21N2S3","21N3S2")
N.plus.phage <- mut.PCoA$points[N.plus.phage.list, ]

#P.no.phage.list <-c("01P3S1","01P4S5","01P5S1","03P3S5","05P3S1",
                    "12P3S4","16P3S2","19P3S4","21P3S4")
#P.no.phage <- mut.PCoA$points[P.no.phage.list, ]
  
P.plus.phage.list <- c("03P2S4","03P5S5","05P2S1","05P2S2","05P4S4","12P2S5",
                       "12P4S2","16P5S2","19P5S2","21P2S4","21P4S2","21P5S1")
P.plus.phage <- mut.PCoA$points[P.plus.phage.list, ]


# Add points

# N.no.phage
#points(N.no.phage[ ,1], N.no.phage[ ,2], pch = 21,
#       cex = 2, col = "blue", bg = "white", lwd = 2)

# N.plus.phage
points(N.plus.phage[ ,1], N.plus.phage[ ,2], pch = 21,
       cex = 2, col = "blue", bg = "lightblue", lwd = 2)

# P.no.phage
#points(P.no.phage[ ,1], P.no.phage[ ,2], pch = 21,
#       cex = 2, col = "red", bg = "white", lwd = 2)

# P.plus.phage
points(P.plus.phage[ ,1], P.plus.phage[ ,2], pch = 21,
       cex = 2, col = "red", bg = "pink", lwd = 2)

```

## summarize by treatment  
Consolidate columns (strains) that have the come from the same treaatment (nutrient and phage)
```{r}
# test for compatability between strain and mutation tables

# strain names from mutation data
breseq.strains <- 
  breseq%>%
  select_if(is.numeric)%>%
  select(-position)%>%
  colnames()

# sum(!(breseq.strains %in% strains$strain_ID)) #0
# sum(!(strains$strain_ID %in% breseq.strains)) #0
# # the lists match!
trtXg <- 
#transpose the table 
mXg %>%
  select(-description)%>%
  pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
  pivot_wider(id_cols = strain_ID, names_from = "gene", values_from = "mutations")%>%
  # add the meta data
  left_join(select(strains,trt, lim, strain_ID), . )%>%
  #summarize treatments
  group_by(trt, lim)%>%
  mutate(n.strains=n())%>%
  group_by(trt, lim,n.strains)%>%
  summarize_if(is_double, sum, na.rm = TRUE)%>%
  # transpose back
  mutate(meta=paste0(lim,".",trt,".","n", n.strains))%>%
  ungroup()%>%
  select(-trt, -lim, -n.strains)%>%
  pivot_longer( cols = c(1:(ncol(.)-1)),names_to = "gene", values_to = "mutations")%>%
  pivot_wider(id_cols = gene, names_from = "meta", values_from = "mutations")


trtXg
```

```{r}

#To normalize mutations I want number of strains
n.strains <- 
  mXg %>%
    select(-description)%>%
    pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
    pivot_wider(id_cols = strain_ID, names_from = "gene", values_from = "mutations")%>%
    # add the meta data
    left_join(select(strains,trt, lim, strain_ID), . )%>%
    #summarize treatments
    group_by(trt, lim)%>%
    mutate(n=n())%>%
    select(trt,lim,n)%>%
    group_by(trt, lim,n)%>%
    summarise()

#normalize
trtXg%>%
  select(-gene)%>%
  sweep(., 2,FUN="/",STATS=n.strains$n)%>%
  #add gene names back
  mutate(gene=trtXg$gene)%>%
  #prepare for plotting
  pivot_longer(cols = colnames(trtXg)[-1], names_to = "trt", values_to = "norm.mut")%>%
  ggplot(aes(x=trt, y=gene,fill=log10(norm.mut)))+
  geom_tile(colour="white",size=0.25)+
  scale_fill_viridis()->p

# ggsave("heatmap.pdf",p, height = 10)

p  


```

# Consolidate mutations by chemostat  

We need to account for repeated samples from the same chemostat and normalize for different number of chemstats per treamtent combination.

## chemostats per treatment
```{r}

strains%>%
  ggplot(aes(x=time))+
  geom_bar(aes(fill=trt), color="black", width=0.5)+
  facet_wrap(~cID, ncol = 2, dir = "v")+
  theme_cowplot()+
  scale_y_continuous(breaks = c(0,1,2))+
  scale_x_continuous(breaks=seq(0,21,7))+
  ggtitle("Strains sequenced in each chemostat")+
  xlab("Time (days)")+
  ylab("# Strains")

```

I think we will need to treat the no-phage sample of each chemostat as that chemostat's ancestor.

## presence-absence of mutations by chemostat.

```{r, message=FALSE}

chemostatXg <-
#transpose the table
mXg %>%
  select(-description)%>%
  pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
  pivot_wider(id_cols = strain_ID, names_from = "gene", values_from = "mutations")%>%
  # add the meta data
  left_join(select(strains,trt, lim,cID, strain_ID), . )%>%
  #summarize treatments
  group_by(trt, lim, cID)%>%
  summarize_if(is_double, sum, na.rm = TRUE)%>%
  # transpose back
  mutate(meta=paste0(lim,".",cID,".",trt))%>%
  ungroup()%>%
  select(-trt, -lim, -cID)%>%
  pivot_longer( cols = c(1:(ncol(.)-1)),names_to = "gene", values_to = "mutations")%>%
  pivot_wider(id_cols = gene, names_from = "meta", values_from = "mutations")#%>%
  # print()


# in a different var I will transform data to presnce absence
chemostatXg.pa <- 
  chemostatXg %>% 
    mutate_at(vars(-gene), ~ replace(., . > 0 | is.na(.), 1))


# aggregate by chemostat
 
chemostatXg.pa%>%
  # transpose
  pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
  mutate(lim=str_split(strain_ID,pattern = "\\.", simplify = T)[,1],
         cID=str_split(strain_ID,pattern = "\\.", simplify = T)[,2],
         trt=str_split(strain_ID,pattern = "\\.", simplify = T)[,3])%>%
  select(-strain_ID)%>%
  #summarize treatments
  group_by(gene,trt ,lim, cID)%>%
    summarize(n.mut=sum(mutations))%>%
  ungroup()%>%
  mutate(n.mut=as.character(n.mut))%>%
    ggplot(aes(x=trt, y=gene,fill=n.mut))+
  geom_tile(colour="white",size=0.25)+
  scale_fill_viridis(discrete = T, direction = -1, option = "magma")+
  facet_grid(.~cID)+
  theme_cowplot() ->p


  ggsave2("mutXchemo_PA.pdf", p,width = 8.5,height = 11)
  
p


```

Intersting observations:  
* gene 1386 has mutations in phage samples of different chemostats.
* Same for 0219, 0102 and several others.
* There are a few mutations that appeear in the non-infected ancestral population and disappear in the evolved infecred. E.g. genes 2307 and hupE in N3.

## numbers to normalize to
i.e. what is the number of independent observations


```{r, include=F}
# #To normalize mutations I want number of chemostats per treatment
# n.strains <- 
#   strains%>%
#   select(trt,lim,cID)%>%
#   filter(!duplicated(interaction(trt,lim,cID)))%>%
#   arrange(cID)%>%
#   group_by(trt, lim)%>%
#     summarise(n=n(), cIDs=paste(cID, collapse=" "))%>%
#   print()
```

```{r, include=F}

chemostatXg <-
#transpose the table
mXg %>%
  select(-description)%>%
  pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
  pivot_wider(id_cols = strain_ID, names_from = "gene", values_from = "mutations")%>%
  # add the meta data
  left_join(select(strains,trt, lim,cID, strain_ID), . )%>%
  #summarize treatments
  group_by(trt, lim, cID)%>%
  summarize_if(is_double, sum, na.rm = TRUE)%>%
  # transpose back
  mutate(meta=paste0(lim,".",cID,".",trt))%>%
  ungroup()%>%
  select(-trt, -lim, -cID)%>%
  pivot_longer( cols = c(1:(ncol(.)-1)),names_to = "gene", values_to = "mutations")%>%
  pivot_wider(id_cols = gene, names_from = "meta", values_from = "mutations")#%>%
  # print()


# in a different var I will transform data to presnce absence
chemostatXg.pa <- 
  chemostatXg %>% 
    mutate_at(vars(-gene), ~ replace(., . > 0 | is.na(.), 1))


# aggregate by treatment
 
chemostatXg.pa%>%
  # transpose
  pivot_longer( cols = c(2:ncol(.)),names_to = "strain_ID", values_to = "mutations")%>%
  mutate(lim=str_split(strain_ID,pattern = "\\.", simplify = T)[,1],
         cID=str_split(strain_ID,pattern = "\\.", simplify = T)[,2],
         trt=str_split(strain_ID,pattern = "\\.", simplify = T)[,3])%>%
  select(-strain_ID)%>%
  #summarize treatments
  group_by(gene,trt ,lim)%>%
    summarize(n.mut=sum(mutations))%>%
  ungroup()%>%
  mutate(n.mut=as.character(n.mut))%>%
    ggplot(aes(x=trt, y=gene,fill=n.mut))+
  geom_tile(colour="white",size=0.25)+
  scale_fill_viridis(discrete = T, direction = -1, option = "magma")+
  facet_grid(.~lim)+
  theme_cowplot()
  



```  

```{r, include=F}
#plot heat map
chemostatXg.pa%>%
  select(-gene)%>%
  sweep(., 2,FUN="/",STATS=n.strains$n)%>%
  #add gene names back
  mutate(gene=trtXg$gene)%>%
  #prepare for plotting
  pivot_longer(cols = colnames(trtXg)[-1], names_to = "trt", values_to = "norm.mut")%>%
  ggplot(aes(x=trt, y=gene,fill=log10(norm.mut)))+
  geom_tile(colour="white",size=0.25)+
  scale_fill_viridis()->p

# ggsave("heatmap.pdf",p, height = 10)

p  
```

